Generalized measurement invariance tests with application to factor analysis

### 2012-02-09 www/rev1/response.txt

Accompanying this letter is our revised manuscript, titled "Generalized
measurement invariance tests with application to factor analysis."  We found the
reviews to be very helpful and believe the paper has improved considerably as a
result.

The main addition to the manuscript is an application of our proposed tests to a
four-group model with mean structure in a study of stereotype threat by
Wicherts, Dolan, and Hessen (2005), as suggested by Reviewer 1.  We elected to
use this real example to supplement, vs replace, the artificial example. While
the real application allows us to illustrate how the tests can be adapted to
specific challenges, the artificial example is more flexible in the illustration
of other general aspects of the tests. Also, it serves as a gateway to the
simulation.  We are willing to streamline this further if you or the reviewers
feel that there is too much redundancy between the artificial and the real
example, however.
README

Furthermore, we have added some theory for the study of local alternatives as
well as details on the use of our methods with categorical V, the ability of our
methods to handle multiple groups models and mean structure, and the
relationship of our methods to mixture models and moderated factor models.

We look forward to receiving feedback on the manuscript.

Sincerely,
Ed Merkle

Editor Comments

@@

Comment E.1. While I share the reviewers generally positive view of your paper,
I find myself in the unusual position of being less enthusiastic about the
present version of the paper than the reviewers seem to be.  One reason for this
is that I think that your method is much less practical than the reviewers do,
chiefly because in practical applications of invariance work, it is unusual to
encounter a case in which V is continuous. The one exception to this situation
is when V is age, but even then, in the context of employment discrimination for
example, legal considerations often demand an arbitrary split on 40 years, or
some other such threshold (which is admittedly arbitrary, but is demanded by the
law).

Response E.1. Though we did not highlight it in the paper, the methods can be
used with categorical V.  We now briefly describe several strategies for
handling ordinal data in the Discussion.  Additionally, we would argue that many
applications of invariance do not use continuous V because methods for handling
continuous V are still emerging (e.g., the moderated factor models discussed by
Reviewer 1).  We believe that our proposed methods will encourage further
invariance research involving continuous V.

@@

Comment E.3. You also did not mention how model identification constraints,
which in the invariance case are sometimes invariance constraints, are
handled.

Response E.3. The proposed tests are based on scores associated with all free
parameters.  Thus, it is straightforward to handle parameters that are
constrained to be equal to one another.  However, this comment may also refer to
the fact that one's choice of identification constraint will influence tests of
other free parameters.  We now briefly describe this issue in the "Use in
Practice" section of the Discussion.

Comment E.4. I think that we need to have a real example, as suggested by
Reviewer one.  This would allow you to answer my second point above, and also
show that your method can give answers that are sensible in context.  With a
real example, you could drop the artificial one (but keep the simulations).
Having a real example would also help to illustrate that there are real
applications in which your method would prove useful.

Response E.4. As discussed in the introduction above, we have added a real
example and hope that it brings out more clearly how the proposed tests can
be adapted to some of the challenges you and the reviewers raised. However,
we have also kept the artificial example because we feel that some of the basic
properties of the tests are easier to illustrate under controlled conditions.
README

@@

Comment 1.3. Similarly, there exist a family of moderation-type factor models
that may serve the same purpose of studying failures of invariance in common
factor models with respect to (unknown) "grouping" variables. For instance,
Dylan Molenaar, Conor Dolan and colleagues have recently worked out CFA models
that were based on Shaun Purcell's work in which certain parameters are
moderated by either observed or unobserved variables. These models too could be
applied to the important problems related to invariance with respect to unknown
groups although they may not work as well as do mixture models and the current
approach when indeed there is some step function that describes the violations
of invariance. The existence of these alternative approaches by no means
diminishes the potential value of the current developments, but it does mean
that the authors should discuss the relation between their approach and other
existing methods.

Response 1.3. We now describe factor analysis models employing moderation in the
"Comparison to other methods" section of the discussion.

Comment 1.4. The authors chose to simulate data only under models that did not
include the mean structure. Since violations of measurement invariance with real
data are most often related to the mean structure (to be sure due to larger
power associated with modeling first-order moments), it may be more realistic
and interesting to have scenarios in which such types of failures of invariance
(called uniform measurement bias) occur.

Response 1.4. Mean parameters were estimated and tested in the simulation
models, though they were unstructured.  Mean structures are often made possible
by constraining parameters across groups, and our simulation focuses on a case
where groups cannot be predefined.  Thus, we have not modified the model in the
simulation but point out more clearly what the 19 parameters in the model are.
In the real data example that we have added, however, the mean structure is
utilized. 
README

@@

Comment 3.3. p. 7 second paragraph: Assuming that the model is correctly
specified is bonus when the second point above is incorporated. I think this
would improve the manuscript.

Response 3.3. Violations of distributional assumptions may still cause problems
with the proposed tests, even if the "structure" of the model (e.g., parameter
constraints, included variables, etc) is correct.  We have modified this section
of text accordingly.

Reviewer #4

Comment 4.1. Page 6/7:	It could be emphasized that a known cutpoint nu means
"observed heterogeneity" and an unknown cutpoint nu means "unobserved
heterogeneity".

Response 4.1. We have now added this detail.

Comment 4.2. Page 9:	The authors should give a reference or a short
explanation of the "tied-down Bessel process". Readers might be unfamiliar
with the process.

Response 4.2. We now note that Andrews (1993) provides results and references
relevant to Bessel processes and also included a forward reference to later
sections. Then, in the "Critical Values & p-values" section, we then state a
brief formal definition of the process (sup_t (t * (1 - t))^{-1} ||B^0(t)||^2)
and point to Hansen (1997) where simulations for critical values are discussed
in detail.

Comment 4.3. Page 10:	To test for changes "along" V means that V is always
continuous (also mentioned in the first paragraph). What happens if V is an
ordered categorical variable? Would that be a special case? Adding a reference
to the discussion section would be helpful.

@@

Response 4.3. We have added some discussion on the use of categorical V.  Also
see Response E.1.

Comment 4.4. Page 19:	The overall k=19 model parameter consists of six factor
loadings, six measurement errors and one covariance between the factors. The
authors refer to the six means of the manifest variables. Usually, in factor
models mean differences of the factors are of research interest. The authors
should give some information why they considered the manifest means instead of
the latent mean differences.

Response 4.4. Factor means can be estimated if one is constraining other model
parameters to be equal across groups.  In our new data example, we do estimate
factor means and use them in the proposed tests.  In our simulation we use a
scenario where no grouping variables exist (i.e., there is only a continuous
V).  In this scenario, it seems difficult to estimate factor means prior to
model estimation because we cannot constrain parameters across groups.  We have
added some extra detail to the simulation section on our use of a single-group
model.

### 2012-02-09 www/paper.Rnw
Addressed all readmes and fixmes in the paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@102 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\acknowledgements{This work was supported by National Science
  Foundation grant SES-1061334. The authors thank Jelte Wicherts, who
  generously shared data for the stereotype threat example.
  Correspondence to Edgar C.\ Merkle, 
  Department of  
  Psychological Sciences, University of Missouri, Columbia,
  MO 65211.
  Email: {\texttt{merklee@missouri.edu}}.}
\shorttitle{Generalized measurement invariance tests}
\rightheader{Generalized measurement invariance tests}

@@

Many expositions of factor analysis utilize the likelihood for the sample
covariance matrix (instead of the likelihood of the individual observations $\bm{x}_i$),
employing a Wishart distribution. However, in that approach the sample covariance
matrix can also be disaggregated to the sum of its individual-specific contributions
leading essentially to a likelihood like \eqref{eq:mvndensity} with multivariate normal
observations $\bm{x}_i$. This situation is also generally encountered in structural
equation models with missing data (e.g., \citeNP{Wot00}).

\readme{maybe like this?}

@@

As mentioned in the theory section, the 
tests give us the flexibility to study hypotheses of partial change.
That is, we have the ability to 
test various subsets of 
parameters.  For example, if we suspected that the verbal factor loadings
lacked measurement invariance, we could test
\begin{equation}
    \label{eq:simh0}
H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
(\lambda_{0, 11}\ \lambda_{0, 21}\ \lambda_{0, 31}),\ i=1,\ldots,n,
\end{equation}
where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
verbal factor loading parameters for student $i$. Thus, here only $k^* = 3$
from the overall $k = 19$~model parameters (including means) are assessed.
\fixme{Explain what the 19 parameters are.}
Alternatively, we can consider all $k^* = k = 19$~parameters, leading to a test
of~\eqref{eq:h0}.  We consider both of these tests below.

@@

\readme{now called ``Application'' as opposed to the artificial ``Example''}

@@

\fixme{Include path diagram and a bit more background so that the paper is more
accessible for readers that are not familiar with the WDH paper.}

To study stereotype threat in a measurement invariance framework, Study~1 of
\citeA{WicDol05} involved 295 high school students
completing three intelligence tests in two between-subjects conditions. 
Conditions were defined by whether or not students received primes about ethnic
stereotypes prior to testing. To study the data, WDH employed a
four-group, one-factor model with the three intelligence tests as manifest
variables. The groups were defined by ethnicity (majority/minority) and by the
experimental manipulation (received/did not receive stereotype prime). Results
indicated that the intelligence tests lacked measurement invariance, with the
minority group receiving stereotype primes being particularly different from the
other three groups on the most difficult intelligence test.  In the
example below, we employ one of the models used by WDH.
Our $V$ variable is participants' grade point averages (GPAs), which were unused in the original analyses.
\fixme{Maybe add: ``\dots as a measure of aptitude.'' Or something along these lines? I'm not
sure what the appropriate psychometric terminology is.}

\subsection{Method}
WDH tested a series of confirmatory factor models for
various types of measurement invariance.  We focus on the model used
in their Step~5b (see pp.~703--704 of their paper), which involved
across-group parameter restrictions on the factor loadings, unique
variances, intercepts, and factor variances.  These parameters were
typically restricted to be equal across groups, though a subset of the
parameters were allowed to be group-specific upon examination of
modification indices.    The model
provided a reasonable fit to the data, as judged by examination of
$\chi^2$, RMSEA (root-mean-square error of approximation), and CFI
(comparative fit index) statistics.  While the model included four
groups as described above, we focus only on the submodel for the minority
group that received stereotype primes.

@@

As mentioned in the theory section, the 
tests give us the flexibility to study hypotheses of partial change. That is, we
have the ability to  test various subsets of  parameters.  For the WDH data, we
first test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i, \text{num}}\ \eta_{i}\ \mu_{i, \text{num}}\ \psi_{i, \text{num}}) = 
      (\lambda_{0, \text{num}}\ \eta_{0}\ \mu_{0, \text{num}}\ \psi_{0, \text{num}}),\
      i = 1,\ldots, n_{\text{MSP}},
\end{equation}
where
$\lambda_{i, \text{num}}$ is the factor loading of student $i$'s intelligence on the numerical scale,
$\eta_{i}$ is student $i$'s factor mean,
$\mu_{i, \text{num}}$ is student $i$'s intercept on the numerical scale, and
$\psi_{i, \text{num}}$ is student $i$'s unique variance on the numerical scale.
The students are ordered according to GPA (i.e.,  in this application the variable $V$),
and the  hypothesis is taken only with respect to the $n_{\text{MSP}}$ individuals in the
``minority, stereotype prime'' group.  Thus, \eqref{eq:exh0} states  that these four group-specific
model parameters are invariant with respect to GPA.

\readme{I have omitted the single parameter hypothesis. Still it would be good to
double-check the estfun method once more, though.}

@@

It is worth mentioning that the proposed tests are not invariant
with respect to the choice of identification constraints. They share this
property with both LM and Wald tests (while LR tests are invariant to
the choice of identification constraints). However, asymptotically
the influence of the parametrization disappears and also in finite
samples it typically does not change the results significantly.
For example, in the stereotype threat application, we assessed the p-values
of all three tests under the constraint that the abstract reasoning
loading is fixed to 1 \fixme{figure ref} and compared them to the p-values
under the constraint that the verbal loading is fixed to 1. All p-values
remained clearly significant and the changes were all smaller than
\Sexpr{format(round(max(abs(p.diffs[1:3,2] - p.diffs[1:3,1])), digits = 5), scientific = FALSE)}.

@@

\readme{slightly rephrased}

@@

\subsection{Categorical Auxiliary Variable}
One issue that was largely unaddressed in this paper involved the use
of categorical $V$ to study measurement invariance.  In this case,
groups are 
already specified in advance, and so traditional methods for fixed
subgroups (i.e., LR, Wald, and LM tests) may suffice.
Furthermore, we can also obtain an LM-type statistic from the
framework developed here. Assume the observations are divided into
$C$ categories $I_1, I_2, \ldots, I_C$. Then, the increment of
the cumulative score process $\Delta_{I_c} {\bm B}(\hat {\bm \theta})$
within each category is just the sum of the corresponding scores.
In somewhat sloppy notation:
\begin{equation}
    \label{eq:catsum}
    \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i \in I_c} {\bm s}(\hat {\bm \theta}; x_i)
%% \\
%%    X^2 & = & \sum_{c = 1}^C n_c^{-1} n^{-1} || \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ||^2
\end{equation}
This results in a $C \times k$ matrix, with one entry for
each category-by-model parameter combination.  We can test
a specific model parameter for invariance by focusing on the associated
column of the $C \times k$ matrix and employing a weighted squared sum of
the entries in the column to obtain a $\chi^2$-distributed statistic with
$(C-1)$ degrees of freedom \cite{HjoKon02}.
Alternatively, to simultaneously test multiple parameters, we can sum
the $\chi^2$ statistics and degrees of freedom for the individual
parameters.  In addition to categorical $V$, this framework may also
be useful for ordinal $V$ (or continuous $V$ with many ties).
Furthermore, in this situation, one could also adapt the statistics
\eqref{eq:dmax}, \eqref{eq:cvm}, and \eqref{eq:maxlm} computed from the same cumulative score
process as usual. The only modification required should be that the statistics
should not depend on the process's values ``within'' a category (or group
of ties). This is easily achieved by taking the maximum (or sum) not over
all observations $i = 1, \dots, n$ but only those $i$ at the ``end'' of a
category.
%% If $i_j$ is the cumulative frequency of category $j$ or lower
%% (with $j = 1, \dots, C$), then one could take the outer maximum in
%% \eqref{} over $i = i_1, i_2, \dots, i_{C-1}$.

\readme{Is this intelligible or do we need more information?}

>>>>

\acknowledgements{This work was supported by National Science
  Foundation grant SES-1061334. The authors thank Jelte Wicherts, who
  generously shared data for the stereotype threat application.
  Correspondence to Edgar C.\ Merkle, 
  Department of  
  Psychological Sciences, University of Missouri, Columbia,
  MO 65211.
  Email: {\texttt{merklee@missouri.edu}}.}
\shorttitle{Generalized measurement invariance tests}
\rightheader{Generalized measurement invariance tests}

@@

Many expositions of factor analysis utilize the likelihood for the sample
covariance matrix (instead of the likelihood of the individual observations $\bm{x}_i$),
employing a Wishart distribution. However, in that approach the sample covariance
matrix can also be disaggregated to the sum of its individual-specific contributions, 
leading essentially to a likelihood like \eqref{eq:mvndensity} with multivariate normal
observations $\bm{x}_i$. This situation is also generally encountered in structural
equation models with missing data (e.g., \citeNP{Wot00}).

@@

As mentioned in the theory section, the 
tests give us the flexibility to study hypotheses of partial change.
That is, we have the ability to 
test various subsets of 
parameters.  For example, if we suspected that the verbal factor loadings
lacked measurement invariance, we could test
\begin{equation}
    \label{eq:simh0}
H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
(\lambda_{0, 11}\ \lambda_{0, 21}\ \lambda_{0, 31}),\ i=1,\ldots,n,
\end{equation}
where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
verbal factor loading parameters for student $i$. Thus, here only $k^* = 3$
from the overall $k = 19$~model parameters are assessed (where the 19 parameters include six factor loadings, six unique variances, six intercepts, and one factor correlation).
Alternatively, we can consider all $k^* = k = 19$~parameters, leading to a test
of~\eqref{eq:h0}.  We consider both of these tests below.

@@

@@

To study stereotype threat in a measurement invariance framework, Study~1 of
\citeA{WicDol05} involved 295 high school students
completing three intelligence tests in two between-subjects conditions. 
Conditions were defined by whether or not students received primes about ethnic
stereotypes prior to testing. To study the data, WDH employed a
four-group, one-factor model with the three intelligence tests as manifest
variables. The groups were defined by ethnicity (majority/minority) and by the
experimental manipulation (received/did not receive stereotype prime). Results
indicated that the intelligence tests lacked measurement invariance, with the
minority group receiving stereotype primes being particularly different from the
other three groups on the most difficult intelligence test.  In the
example below, we employ one of the models used by WDH.
Our $V$ is participants' aptitudes, as measured by their grade point averages (GPAs).  The GPAs were unused in the original analyses.

\subsection{Method}
WDH tested a series of confirmatory factor models for
various types of measurement invariance.  We focus on the model used
in their Step~5b (see pp.~703--704 of their paper), which involved
across-group restrictions on the factor loadings, unique
variances, intercepts, and factor variances.  These parameters were
typically restricted to be equal across groups, though a subset of the
parameters were allowed to be group-specific upon examination of
modification indices.    The model
provided a reasonable fit to the data, as judged by examination of
$\chi^2$, RMSEA (root-mean-square error of approximation), and CFI
(comparative fit index) statistics.  While the model included four
groups as described above, we focus only on the submodel for the minority
group that received stereotype primes, which is pictured in Figure~\ref{fig:wpath}.  In the figure, paths with dashed lines and red parameters signify group-specific parameters.  These include the factor mean $\eta$, numerical factor loading $\lambda_{\text{num}}$, numerical intercept $\mu_{\text{num}}$, and numerical unique variance $\psi_{\text{num}}$.  Other parameters (not displayed) were restricted to be equal across all four groups, with the exception of the factor variance.  This parameter was restricted to be equal within both minority groups, and separately within both majority groups.

\begin{figure}
\caption{Path diagram representing the submodel for the ``minority, stereotype'' group in model 5b of WDH.  Dashed paths and red parameters are group specific.  Other parameters (not displayed) were restricted to be equal across all four groups, with the factor variance only being restricted across the two minority groups.}
\label{fig:wpath}
\includegraphics[height=3.5in]{wicherts_msp.pdf}
\end{figure}

@@

As mentioned in the theory section, the 
tests give us the flexibility to study hypotheses of partial change. That is, we
have the ability to  test various subsets of  parameters.  For the WDH data, we
first test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i, \text{num}}\ \eta_{i}\ \mu_{i, \text{num}}\ \psi_{i, \text{num}}) = 
      (\lambda_{0, \text{num}}\ \eta_{0}\ \mu_{0, \text{num}}\ \psi_{0, \text{num}}),\
      i = 1,\ldots, n_{\text{MSP}},
\end{equation}
where
$\lambda_{i, \text{num}}$ is the factor loading on the numerical scale,
$\eta_{i}$ is student $i$'s factor mean,
$\mu_{i, \text{num}}$ is student $i$'s intercept on the numerical scale, and
$\psi_{i, \text{num}}$ is student $i$'s unique variance on the numerical scale.
The students are ordered according to GPA (i.e.,  in this application the variable $V$),
and the  hypothesis is taken only with respect to the $n_{\text{MSP}}$ individuals in the
``minority, stereotype prime'' group.  Thus, \eqref{eq:exh0} states  that these four group-specific
model parameters are invariant with respect to GPA.

@@

It is worth mentioning that the proposed tests are not invariant
with respect to the choice of identification constraints. They share this
property with both LM and Wald tests (while LR tests are invariant to
the choice of identification constraints). However, asymptotically
the influence of the parametrization disappears and also in finite
samples it typically does not change the results significantly.
For example, in the stereotype threat application, we assessed the p-values
of all three tests under the constraint that the abstract reasoning
loading is fixed to 1 (see Figure~\ref{fig:wpath}) and compared them to the p-values
under the constraint that the verbal loading is fixed to 1. All p-values
remained clearly significant and the changes were all smaller than
\Sexpr{format(round(max(abs(p.diffs[1:3,2] - p.diffs[1:3,1])), digits = 5), scientific = FALSE)}.

@@

@@

\subsection{Categorical Auxiliary Variable}
One issue that was largely unaddressed in this paper involved the use
of categorical $V$ to study measurement invariance.  In this case,
groups are 
already specified in advance, and so traditional methods for fixed
subgroups (i.e., LR, Wald, and LM tests) may suffice.
Furthermore, we can also obtain an LM-type statistic from the
framework developed here. Assume the observations are divided into
$C$ categories $I_1, I_2, \ldots, I_C$. Then, the increment of
the cumulative score process $\Delta_{I_c} {\bm B}(\hat {\bm \theta})$
within each category is just the sum of the corresponding scores.
In somewhat sloppy notation:
\begin{equation}
    \label{eq:catsum}
    \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i \in I_c} {\bm s}(\hat {\bm \theta}; x_i)
%% \\
%%    X^2 & = & \sum_{c = 1}^C n_c^{-1} n^{-1} || \Delta_{I_c} {\bm B}(\hat {\bm \theta}) ||^2
\end{equation}
This results in a $C \times k$ matrix, with one entry for
each category-by-model parameter combination.  We can test
a specific model parameter for invariance by focusing on the associated
column of the $C \times k$ matrix and employing a weighted squared sum of
the entries in the column to obtain a $\chi^2$-distributed statistic with
$(C-1)$ degrees of freedom \cite{HjoKon02}.
Alternatively, to simultaneously test multiple parameters, we can sum
the $\chi^2$ statistics and degrees of freedom for the individual
parameters.  In addition to categorical $V$, this framework may 
be useful for ordinal $V$ (or continuous $V$ with many ties).
In this situation, one could also adapt the statistics
\eqref{eq:dmax}, \eqref{eq:cvm}, and \eqref{eq:maxlm} computed from the same cumulative score
process as usual. The only required modification is that the statistics
should not depend on the process's values ``within'' a category (or group
of ties). This is easily achieved by taking the maximum (or sum) not over
all observations $i = 1, \dots, n$, but over only those $i$ at the ``end'' of a
category.
%% If $i_j$ is the cumulative frequency of category $j$ or lower
%% (with $j = 1, \dots, C$), then one could take the outer maximum in
%% \eqref{} over $i = i_1, i_2, \dots, i_{C-1}$.
